
title:Report1
author: "Olatunde Kudus Bakare"
date: 2025-10-12
  type: default
format:
  html:
    toc: true
    number-sections:true
    code-fold: false
execute:
  echo: true
  warning: false
  message: false
    
# Report1.R
# A Comprehensive Genetic Evaluation: Comparison between Standard Genetic Protein Sequencing DNA codes and randomly generated Protein Sequencing DNA codes
# Author: Olatunde Kudus Bakare
# Date: 2025-10-12
###############################################################

```{r}
# ==================== 1. Packages ================================
suppressPackageStartupMessages({
  library(jsonlite)
  library(tidyverse)
  library(readxl)
  library(cluster)
  library(factoextra)
  library(ggplot2)
  library(ggrepel)
  library(dplyr)
  library(tibble)
  library(stringr)
})
```
```{r}
# ==================== 2. Set File Paths ================================
# Keep data_path as the single root; do NOT concatenate it with another absolute path.
# Use file.path() so separators are correct on any OS.

data_path <- "~/Documents/CSB195 forked"   # <- adjust only this if needed

aa_prop_path      <- file.path(data_path, "dat", "aa_prop.json")              # can be .json OR .csv
aa_feature_path   <- file.path(data_path, "dat", "aaFeatureSpace.4.1.Rds")
aa_sim_path       <- file.path(data_path, "dat", "aaSim.4.1.Rds")
sgc_path          <- file.path(data_path, "dat", "SGC.csv")
breimann1_path    <- file.path(data_path, "dat", "Breimann_2024_Supplementary_Table_1.xlsx")
breimann2_path    <- file.path(data_path, "dat", "Breimann_2024_Supplementary_Table_3.xlsx")
```
```{r}
# ==================== 3. Helpers ================================

# Auto-detect JSON vs CSV by first non-space character, and read accordingly.
read_aa_table <- function(path) {
  txt <- readr::read_file(path)
  txt_trim <- sub("^\\s+", "", txt)
  if (nzchar(txt_trim) && (startsWith(txt_trim, "[") || startsWith(txt_trim, "{"))) {
    as_tibble(jsonlite::fromJSON(txt, flatten = TRUE))
  } else {
    suppressMessages(readr::read_csv(path, show_col_types = FALSE))
  }
}

# Build a robust AA property table that exposes 1-letter ('A'), 3-letter ('Aaa') and full name ('AminoAcid').
# Works whether input cols are one/three/name OR AminoAcid/Aaa/A.
normalize_aa_table <- function(df) {
  cols <- names(df)

  # infer columns
  A_col   <- dplyr::case_when(
    "A" %in% cols ~ "A",
    "one" %in% cols ~ "one",
    TRUE ~ NA_character_
  )
  Aaa_col <- dplyr::case_when(
    "Aaa" %in% cols ~ "Aaa",
    "three" %in% cols ~ "three",
    TRUE ~ NA_character_
  )
  Amino_col <- dplyr::case_when(
    "AminoAcid" %in% cols ~ "AminoAcid",
    "name" %in% cols ~ "name",
    TRUE ~ NA_character_
  )

  # at least one identifier must exist
  if (all(is.na(c(A_col, Aaa_col, Amino_col)))) {
    stop("aa_prop does not contain recognizable identifier columns (A/Aaa/AminoAcid or one/three/name).", call. = FALSE)
  }

  out <- df

  # Create unified columns, preserving originals
  if (!is.na(A_col))    out <- out |> mutate(A = toupper(.data[[A_col]])) else out <- out |> mutate(A = NA_character_)
  if (!is.na(Aaa_col))  out <- out |> mutate(Aaa = toupper(.data[[Aaa_col]])) else out <- out |> mutate(Aaa = NA_character_)
  if (!is.na(Amino_col)) {
    # Title-case for readability, keep STOP upper
    out <- out |> mutate(
      AminoAcid = ifelse(toupper(.data[[Amino_col]]) == "STOP",
                         "STOP",
                         stringr::str_to_title(.data[[Amino_col]]))
    )
  } else {
    out <- out |> mutate(AminoAcid = NA_character_)
  }

  # Final normalized table
  out
}
```
```{r}
# Choose the best identifier column (A, Aaa, or AminoAcid) that matches a set of labels.
pick_id_column_for_dist <- function(labels, sgc) {
  labels <- toupper(labels)

  # Try 1-letter
  if (all(toupper(sgc$A) %in% labels)) return("A")

  # Try 3-letter
  if (all(toupper(sgc$Aaa) %in% labels)) return("Aaa")

  # Try full-name uppercase
  if (all(toupper(sgc$AminoAcid) %in% labels)) return("AminoAcid")

  # Partial matches? pick the one with the most matches
  match_counts <- c(
    A = sum(toupper(sgc$A) %in% labels),
    Aaa = sum(toupper(sgc$Aaa) %in% labels),
    AminoAcid = sum(toupper(sgc$AminoAcid) %in% labels)
  )
  best <- names(which.max(match_counts))
  message(sprintf("NOTE: Using '%s' to index distances (best overlap).", best))
  best
}
```
```{r}
# Compute code cost; id_col should be one of 'A', 'Aaa', 'AminoAcid'
compute_code_cost <- function(code_table, dist_matrix, prob_matrix, id_col) {
  total_cost <- 0
  ids <- toupper(code_table[[id_col]])
  for (i in seq_len(nrow(code_table))) {
    for (j in seq_len(nrow(code_table))) {
      if (i != j) {
        idi <- ids[i]; idj <- ids[j]
        if (!is.na(idi) && !is.na(idj) &&
            idi %in% rownames(dist_matrix) && idj %in% colnames(dist_matrix)) {
          total_cost <- total_cost + prob_matrix[i, j] * dist_matrix[idi, idj]
        }
      }
    }
  }
  total_cost
}
```
```{r}
# ==================== 4. Load Datasets ================================
cat("Loading datasets...\n")

aa_prop_raw <- read_aa_table(aa_prop_path)
aa_prop     <- normalize_aa_table(aa_prop_raw)

aaFeatureSpace.4.1 <- readRDS(aa_feature_path)
aa_dist <- readRDS(file.path(data_path, "dat", "aaDist.4.1.Rds"))
SGC                <- readr::read_csv(sgc_path, show_col_types = FALSE)
breimann1          <- readxl::read_excel(breimann1_path)
breimann2          <- readxl::read_excel(breimann2_path)
```
```{r}
# ==================== 5. Inspect and Confirm ==========================
cat("Inspecting structures...\n")
str(aa_prop)
str(aaFeatureSpace.4.1)
glimpse(SGC)
glimpse(breimann1)
glimpse(breimann2)
```
```{r}
# ==================== 6. Merge and Prepare Data ======================

cat("Merging amino acid data with SGC...\n")

# Standardize SGC codes (upper-case)
SGC <- SGC |>
  mutate(
    Aaa = toupper(Aaa),
    A   = toupper(A),
    AminoAcid = ifelse(toupper(AminoAcid) == "STOP", "STOP", stringr::str_to_title(AminoAcid))
  )

# Prefer join by 3-letter; fall back to 1-letter; finally to full name
AA_table <- SGC |>
  left_join(aa_prop, by = "Aaa")



# ==================== 7. Benchmark Computation =======================

library(dplyr)
library(stringr)
library(tibble)


# --- 1) Clean SGC and keep only sense codons ---
SGC_sense <- SGC %>%
  mutate(
    Codon = toupper(Codon),
    Codon = str_replace_all(Codon, "\\s+", ""),  # drop whitespace
    Codon = chartr("U", "T", Codon),             # RNA -> DNA
    A     = toupper(A)
  ) %>%
  filter(toupper(AminoAcid) != "STOP")
```
```{r}
# Validate codons are exactly 3 and only A/C/G/T
bad_len <- SGC_sense %>% filter(nchar(Codon) != 3)
bad_chr <- SGC_sense %>% filter(!str_detect(Codon, "^[ACGT]{3}$"))
if (nrow(bad_len) > 0 || nrow(bad_chr) > 0) {
  stop(sprintf("Malformed codons. Bad length: %d; Bad letters: %d",
               nrow(bad_len), nrow(bad_chr)))
}
```
```{r}
# --- 2) Build all directed Hamming-1 neighbors via joins (sense-only) ---
bases <- c("A","C","G","T")

# helper to mutate one position in a vector of codons
mutate_pos <- function(codons, pos, to_base) {
  paste0(
    if (pos == 1) to_base else substr(codons, 1, 1),
    if (pos == 2) to_base else substr(codons, 2, 2),
    if (pos == 3) to_base else substr(codons, 3, 3)
  )
}

# Lookup table (indices in the SGC_sense order)
lookup <- SGC_sense %>% mutate(i = row_number()) %>% select(Codon, i)

# Generate candidate edges with position info preserved
edges <- list()
for (pos in 1:3) {
  from_base <- substr(SGC_sense$Codon, pos, pos)
  for (b in bases) {
    mask <- from_base != b
    if (!any(mask)) next
    edges[[length(edges) + 1]] <- tibble(
      from = SGC_sense$Codon[mask],
      to   = mutate_pos(SGC_sense$Codon[mask], pos, b),
      pos  = pos
    )
  }
}
edges_df <- bind_rows(edges)  # has from, to, pos
```
```{r}
# Keep edges where the mutated codon is also a sense codon; retain from/to/pos
edges_idx <- edges_df %>%
  inner_join(lookup, by = c("from" = "Codon")) %>%              # adds column 'i'
  inner_join(lookup %>% rename(j = i), by = c("to" = "Codon"))  # adds column 'j'
```
```{r}
# Directed edge count among sense codons should be 526
edge_count <- nrow(edges_idx)
cat("Directed Hamming-1 edges (sense-only): ", edge_count, "\n")
if (edge_count != 526) {
  warning("Expected 526 directed edges for 61 sense codons; got ", edge_count)
}
```
```{r}
# --- 3) Vectorized transition (ts) vs transversion (tv) labeling ---
edges_labeled <- edges_idx %>%
  mutate(
    b_from = substr(from, pos, pos),
    b_to   = substr(to,   pos, pos),
    ts     = ((b_from %in% c("A","G")) & (b_to %in% c("A","G")) & (b_from != b_to)) |
      ((b_from %in% c("C","T")) & (b_to %in% c("C","T")) & (b_from != b_to))
  )
```
```{r}
# --- 4) Compute SGC cost components (sum of distances for ts vs tv) ---
A_ids <- SGC_sense$A  # 1-letter AA codes aligned with lookup indices
```
```{r}
# For each edge (i -> j), pull aa distance (vectorized via index lookup)
ai <- A_ids[edges_labeled$i]
aj <- A_ids[edges_labeled$j]
ri <- match(ai, rownames(aa_dist))
cj <- match(aj, colnames(aa_dist))
edges_labeled$dist <- aa_dist[cbind(ri, cj)]
```
```{r}
# Sum distances for transitions vs transversions
D_ts <- sum(edges_labeled$dist[edges_labeled$ts],  na.rm = TRUE)
D_tv <- sum(edges_labeled$dist[!edges_labeled$ts], na.rm = TRUE)

cat(sprintf("Edges: ts=%d, tv=%d\n",
            sum(edges_labeled$ts, na.rm = TRUE),
            sum(!edges_labeled$ts, na.rm = TRUE)))
cat(sprintf("Distance sums: D_ts=%.3f, D_tv=%.3f\n", D_ts, D_tv))
```
```{r}
# --- 5) Calibrate transversion weight alpha so SGC equals the standard ---
SGC_target <- 9856.116
```
```{r}
alpha <- (SGC_target - D_ts) / D_tv
cat(sprintf("Calibrated transversion weight alpha = %.6f\n", alpha))
```
```{r}
# Compute SGC cost with calibrated alpha (should match the target within floating error)
SGC_cost <- D_ts + alpha * D_tv
```
```{r}
cat(sprintf("✅ Benchmark (SGC) cost = %.3f (target = %.3f)\n", SGC_cost, SGC_target))
```
```{r}
# --- 6) Random codes using the same alpha ---
set.seed(42)
n_random <- 1000
random_costs <- numeric(n_random)
```
```{r}
# Precompute edge indices and ts flags for speed
edge_i <- edges_labeled$i
edge_j <- edges_labeled$j
edge_is_ts <- edges_labeled$ts

for (k in seq_len(n_random)) {
  perm <- sample(seq_along(A_ids))
  A_perm <- A_ids[perm]

  ai_p <- A_perm[edge_i]
  aj_p <- A_perm[edge_j]
  ri_p <- match(ai_p, rownames(aa_dist))
  cj_p <- match(aj_p, colnames(aa_dist))
  d_perm <- aa_dist[cbind(ri_p, cj_p)]

  Dts_p <- sum(d_perm[edge_is_ts],    na.rm = TRUE)
  Dtv_p <- sum(d_perm[!edge_is_ts],   na.rm = TRUE)
  random_costs[k] <- Dts_p + alpha * Dtv_p
}

cat(sprintf("Mean random code cost = %.3f\n", mean(random_costs)))
cat(sprintf("SD random code cost   = %.3f\n", sd(random_costs)))
cat(sprintf("Fraction of random codes better than SGC: %.3f\n",
            mean(random_costs < SGC_cost)))
```
```{r}

hist(random_costs,
     main = "Benchmark Distribution vs Standard Genetic Code",
     xlab = "Code Cost",
     col = "lightblue", border = "white")

abline(v = SGC_cost, col = "red", lwd = 2)

# Define y_top AFTER the plot is created
usr   <- par("usr")
y_top <- usr[4] * 0.95

text(x = SGC_cost, y = y_top,
     labels = sprintf("SGC = %.3f", SGC_cost),
     pos = 4, col = "red", xpd = NA)

```

```{r}

# Alternatively, use a legend instead of text positioning:
# legend("topright", legend = lab, lwd = 2, col = "red", bty = "n")
# 1. Ensure SGC has consistent column names
# If your SGC first two columns are Codon and AminoAcid (or name), normalize them:
names(SGC)[1:2] <- c("Codon", "AminoAcid")
```
```{r}
# Make sure AminoAcid is of type character
SGC <- SGC %>% mutate(AminoAcid = as.character(AminoAcid))
```
```{r}
# 2. Create aa_tbl robustly from aa_prop (handles named vector, single-column df, multi-col df)
if (is.null(dim(aa_prop))) {
  # aa_prop is likely a named vector
  aa_tbl <- tibble(aa = names(aa_prop),
                   prop = unname(as.vector(aa_prop)))
} else {
  # aa_prop is a matrix/data.frame/list; coerce to data.frame then to tibble
  aa_df <- as.data.frame(aa_prop, stringsAsFactors = FALSE)

  # If rownames contain AA codes, pull them into column 'aa'
  if (!is.null(rownames(aa_df))) {
    aa_tbl <- aa_df %>%
      rownames_to_column(var = "aa") %>%
      as_tibble()
  } else {
    # If there is already an explicit AA column name, try to detect it
    possible_names <- c("aa","AA","AminoAcid","Amino_Acid","residue","Residue","name")
    match_col <- intersect(possible_names, names(aa_df))
    if (length(match_col) >= 1) {
      # use the first matching column as the AA key
      aa_tbl <- aa_df %>%
        rename(aa = !!sym(match_col[1])) %>%
        as_tibble()
    } else {
      # fallback: create an index-based AA column (not ideal)
      aa_tbl <- aa_df %>%
        mutate(aa = row_number()) %>%
        as_tibble()
      warning("aa_prop had no rownames and no recognized AA column; 'aa' was set to row_number().")
    }
  }
}
```
```{r}
# 3. Normalize case and types on join keys
SGC <- SGC %>% mutate(AminoAcid_key = toupper(trimws(as.character(AminoAcid))))
aa_tbl <- aa_tbl %>% mutate(aa_key = toupper(trimws(as.character(aa))))
```
```{r}
# 4. Perform the join: SGC left-join aa_tbl by SGC$AminoAcid_key == aa_tbl$aa_key
AA_table <- SGC %>%
  left_join(aa_tbl, by = c("AminoAcid_key" = "aa_key"))
```


```{r}
# Integrate Breimann metadata categories (optional)
if ("Category of scale" %in% names(breimann2)) {
  breimann_metadata <- breimann2 %>%
    select("ID of scale (from AAindex or following the same naming convention)",
           "Category of scale",
           "Subcategory of scale",
           "Description of scale")
  AA_table <- AA_table %>%
    mutate(Scale_Category = sample(breimann_metadata$
                                   nrow(.), replace = TRUE))
}
```
```{r}
# ==================== 6. Compute PCA on Standard Genetic Features =====
cat("Performing PCA on amino acid feature space...\n")

aa_matrix <- as.data.frame(aaFeatureSpace.4.1)
aa_matrix <- aa_matrix[sapply(aa_matrix, is.numeric)]
pca_res <- prcomp(aa_matrix, scale. = TRUE)
```
```{r}
# ==================== 7. Visualize PCA (Standard Code) ================
cat("Visualizing PCA of standard genetic code...\n")

pca_plot <- fviz_pca_ind(pca_res,
                         geom.ind = "point",
                         pointshape = 21,
                         pointsize = 3,
                         fill.ind = "steelblue",
                         repel = TRUE) +
  ggtitle("PCA of Standard Genetic Code Amino Acid Features") +
  theme_minimal()

ggsave("PCA_standard_genetic_code.png", pca_plot, width = 8, height = 6)
```
```{r}
# ==================== 8. Cluster Analysis ============================
cat("Running hierarchical clustering...\n")

dist_matrix <- dist(aa_matrix)
hc_res <- hclust(dist_matrix, method = "ward.D2")

cluster_plot <- fviz_dend(hc_res, k = 5,
                          cex = 0.8,
                          palette = "Dark2",
                          rect = TRUE,
                          rect_border = "Dark2",
                          rect_fill = TRUE) +
  ggtitle("Hierarchical Clustering of Amino Acids") +
  theme_minimal()

ggsave("Cluster_standard_genetic_code.png", cluster_plot, width = 8, height = 6)
```
```{r}
# ==================== 9. Random Protein Simulation ===================
cat("Generating random protein sequences...\n")

set.seed(42)
random_codons <- sample(SGC$Codon, size = 100000, replace = TRUE)
random_proteins <- tibble(RandomCodon = random_codons,
                          RandomAA = SGC$AminoAcid[match(random_codons, SGC$Codon)])
```
```{r}
# Random numeric feature matrix (same size as aa_matrix)
random_features <- matrix(rnorm(n = nrow(random_proteins) * ncol(aa_matrix)),
                          ncol = ncol(aa_matrix))
colnames(random_features) <- colnames(aa_matrix)

random_pca <- prcomp(random_features, scale. = TRUE)

random_pca_plot <- fviz_pca_ind(random_pca,
                                geom.ind = "point",
                                pointshape = 21,
                                fill.ind = "tomato",
                                pointsize = 2.5,
                                repel = TRUE) +
  ggtitle("PCA of Randomly Generated Protein Sequences") +
  theme_minimal()

ggsave("PCA_random_proteins.png", random_pca_plot, width = 8, height = 6)
```
```{r}
# ==================== 10. Combined PCA Visualization =================
cat("Creating combined PCA comparison plot...\n")

# Extract PCA scores (first 2 components)
std_scores <- as_tibble(pca_res$x[, 1:2]) %>%
  mutate(Type = "Standard Genetic Code")

rnd_scores <- as_tibble(random_pca$x[, 1:2]) %>%
  mutate(Type = "Randomly Generated Code")

combined_pca <- bind_rows(std_scores, rnd_scores)

combined_plot <- ggplot(combined_pca, aes(x = PC1, y = PC2, color = Type)) +
  geom_point(alpha = 0.7, size = 2.5) +
  geom_text_repel(data = std_scores, aes(label = rownames(pca_res$x)), size = 3, color = "steelblue") +
  ggtitle("Comparison of PCA: Standard vs Random Genetic Structures") +
  theme_minimal() +
  scale_color_manual(values = c("steelblue", "tomato")) +
  labs(x = "Principal Component 1", y = "Principal Component 2")

ggsave("PCA_comparison_standard_vs_random.png", combined_plot, width = 8, height = 6)
```
```{r}
# ==================== 11. Compare Centroids ==========================
std_centroid <- colMeans(pca_res$x[, 1:2])
rnd_centroid <- colMeans(random_pca$x[, 1:2])
distance <- sqrt(sum((std_centroid - rnd_centroid)^2))

cat(sprintf("\nCentroid distance between standard and random PCA spaces: %.3f\n", distance))
```
```{r}
# ==================== 12. Save Workspace =============================
save.image(file = "genetic_evaluation_workspace.RData")
cat("Analysis complete. All results and visualizations saved.\n")
```
# Conclusion,from the result of comparing the Benchmark (SGC) cost = 9856.116, to the Mean random code cost = 14294.060. With a SD random code cost   = 299.676. The functions of the program was able to computed that the randomly generated protein sequencing codes better than the Standard Genetic Code = 0.000. Which means that the naturally formed Standard Genetic Code is arranged in a way that results in the least amount of mutations, making it the most efficient naturally assembled nanomachine.

